Inference using context-dependent models of mutation

### 2017-04-26 writeups/context-dependence.tex
rearranged writeup

It not uncommon to encounter models
consisting of a lattice of sites,
each site taking one of a finite collection of possible states,
and whose stochastic, temporal evolution is Markov and governed by a set of local rules.
For instance, at each site may sit
a nucelotide whose mutation rate depends on physical properties determined by the nearby DNA sequence;
a cell whose infection status depends on the state of its neighbors;
or a particle whose spin is perturbed by external noise to a state whose distribution depends on the local energy configuration.
Practical use of such models often requires
inferring transition rules
based on observations of the system at several time points,
or of several states evolved along a tree from a single starting point.
In this paper,
we propose one solution to this problem,
motivated by the problem of inferring context-dependent mutation rates from diverged nucleotide sequences.

The general class of models we consider are known in the probability literature as
interacting particle systems \citep{liggett2005ips} with neighborhood structure
-- continuous-time Markov chains on lattice-indexed collections of states
whose transition probabilities are \emph{local}
in the sense that any instantaneous change only affects a small number of nearby sites,
and the rates of such instantaneous changes depend only on the states in some bounded neighborhood of those sites to change.
As finite-state Markov chains,
transition probabilities are in principle simply expressable as a matrix exponential,
but this is impractical because the size of the matrix is equal to the number of possible configurations.
\todo{I suggest pulling this sentence together with your "In this paper" sentence to form a separate paragraph.}
As an alternative, we show that the conditional likelihood of local pattern count statistics
can be well-approximated by marginalizing over a finite amount of surrounding context.

@@

%%%%%%% %%%%%%%%%%
\section{Notation and Examples}

Each example above describes \todo{Has a formulation as? Glauber can be greater than 1 D} a 1-dimensional grid of sites, each of which can take one of a finite set of states,
and that switch randomly between states according to a local set of rules.
Concretely, we suppose that we observe a finite collection of these sites at only a few times,
and that the dynamics are Markov:
with the set of possible states by $\calS$,
and $X_i(t)$ the state of site $i$ at time $t$,
we assume that $\{X(t)\}_{t \ge 0}$ is a Markov process on sequences of $n$ states $\calS^n$
for which the probability a given site changes state in a small amount of time
depends only on the sequence of states at nearby sites.

@@

The dynamics of our stochastic process are determined by the set of allowed transitions and associated rates:
given two patterns $u$ and $v$ of common length $h$,
saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that
\[
    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
Such a rule is defined by its ``transition triple'' $(\mu,u,v)$.
The set of all transition triples, $\calT$, determines the time evolution of this process.

\begin{example}[TASEP]
  The \emph{Totally Asymmetric Simple Exclusion Process} labels each site as ``empty'' or ``occupied'' (0 or 1, respectively),
  and says that each occupied site, independently at rate $\lambda$, checks the site to the right to see if it is empty,
  and if it is, moves there.
  This process therefore has only one transition triple \todo{Not sure if reviewers will like having a different notation for triples so soon after you defined the parens notation}:

@@

\begin{example}[1-D Ising model with Glauber dynamics]
  In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
  imagined as magnetic dipoles,
  and the energy associated with a given state $x$ is $H(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
  Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
  The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
  The following ``Glauber dynamics'' preserve the stationary distribution:
  each site, independently at rate $\lambda$,
  forgets its spin,
  and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration.
  Ignoring transitions that don't change the state,
  if patterns $u$ and $v$ differ at only one site then $u$ changes to $v$ at rate $1/(1+\exp(H(v)-H(u)))$ \todo{Shouldn't there be a $\lambda$ up top?}, so

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm.''
.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$ per site,
and then resolving each possible change in temporal order.
The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
Note that since mutation is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
so the maximization over $i$ is unnecessary.
Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
so that for each site $i$,
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a draw from a rate $\mu_*$ Poisson process on $[0,t]$.
These processes are independent across sites.
Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $c_{ij} = \min_{k,h} \{ c_{kh} : c_{kh}>s \}$.
At $c_{ij}$,
a new state is then chosen at position $i$
with probability proportional to the corresponding rate, i.e.\
\erick{Previously, the $j$ in $\mu^j$ was used to index the possible patterns. Now it's indexing possible changes at a site.}
\[
q_j = \begin{cases}
  \mu_j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
is the probability that $x_i^{(|u^j|)}$ is replaced with $v^j$.
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$ gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0\ge 0$.)

@@

@@

%%%%
\subsection{Summary statistics}

%%%%
\subsection*{Proof of the approximation}

The choice to treat long sequences as cyclical is generally of little importance.
However, we have defined in \eqref{eqn:G_defn} the basic building block of our approximation, $G(m)$,
to be the transition matrix \emph{for a circular sequence} of length $m$, where $m$ may be short.
This seems strange -- but, what else would we do at the edges?
We want to use $G(\ell+m+r)$ to compute the chance that a given $(\ell+m+r)$-sequence ends up with another given $m$-sequence in the middle;
and if $\ell$ and $r$ are big enough relative to $t$, this will be a good approximation regardless of what we do at the edges.
One answer would be to not allow patterns hanging off of either end to match,
resulting in a slight underestimate of the chance of change.
A better approximation would be to use the marginal frequencies in some way to average over surroudning sequence
-- but, this requires marginal frequencies to be known, and these are not stationary.
However, the leftmost site in a randomly chosen window is a sample from the marginal distribution,
and if sites separated by distance $\ell+m+r$ are sufficiently uncorrelated,
wrapping the sequence around is a good approximation to averaging over marginal frequencies in the surrounding sequence.
It is easy to imagine situations where this would not be the best solution --
but again, if $\ell$ and $r$ are large enough, it won't matter much.

@@

\subsection*{Algorithm}

%%%%%%
\section{Bayesian things}

So, suppose we've got a list of patterns $(x_i,y_i)$
and we're happy with the idea that they're independent.
Recall that
\[
\E\left[ N(x,y) \mid N(x) \right] = N(x) p_{m,\ell}(x,y) ,
\]
Parameters $\theta$ enter implicitly through $p_{m,\ell}$.

In the case that we observe the initial sequence, $N(x,y)$ are binomial with sample size $N(x)$.
and so the likelihood is
\[
\prod_x N(x)! \prod_y \frac{ p_{m,\ell}(x,y)^{N(x,y)} }{ N(x,y)! } ,
\]
and so the negative log likelihood is, up to a constant,
\[
\sum_{x,y} N(x,y) \log p_{m,\ell}(x,y) .
\]

\emph{The following is copied over from work on the abphylo project.}

%%%%%%%
\section{Simulation results}

The method works well to find point estimates of the parameters.
Obtaining posterior density estimates using the approximation likelihood function on a subset of patterns
works in some cases;
in other cases the likelihood surface is still too peaked, underestimating the uncertainty.

@@

\section{Applications}

graham:
\begin{itemize}

@@

sergey:
\begin{itemize}

@@

@@

%%%%%%%%
\section{What next}

@@

  \item Allow more than single-base distribution at the root.

  \item What's up with the mixing problem in root frequencies?

  \item See if counting in nonoverlapping windows fixes the peakiness problem with TASEP. \emph{Yes, it does.}

  \item See what's feasible with a larger alphabet -- 60 codons -- and more parameters.

  \item Do a three-taxa tree.

@@

>>>>

It not uncommon to encounter models
consisting of a lattice of sites,
each site taking one of a finite collection of possible states,
and whose stochastic, temporal evolution is Markov and governed by a set of local rules.
For instance, at each site may sit
a nucelotide whose mutation rate depends on physical properties determined by the nearby DNA sequence;
a cell whose infection status depends on the state of its neighbors;
or a particle whose spin is perturbed by external noise to a state whose distribution depends on the local energy configuration.
Practical use of such models often requires
inferring transition rules
based on observations of the system at several time points,
or of several states evolved along a tree from a single starting point.

The general class of models we consider are known in the probability literature as
interacting particle systems \citep{liggett2005ips} with neighborhood structure
-- continuous-time Markov chains on lattice-indexed collections of states
whose transition probabilities are \emph{local}
in the sense that any instantaneous change only affects a small number of nearby sites,
and the rates of such instantaneous changes depend only on the states in some bounded neighborhood of those sites to change.
As finite-state Markov chains,
transition probabilities are in principle simply expressable as a matrix exponential,
but this is impractical because the size of the matrix is equal to the number of possible configurations.

In this paper, we propose a solution to this problem,
showing that the conditional likelihood of local pattern count statistics
can be well-approximated by marginalizing over a finite amount of surrounding context.
Motivated by the problem of inferring context-dependent mutation rates from diverged nucleotide sequences,
we extend the inference framework to observations on trees.
\peter{something about efficient computation and sparse matrices}

@@

%%%%%%% %%%%%%%%%%
\section{A general context-dependent model}

Consider a 1-dimensional grid of sites, each of which can take one of a finite set of states,
and that switch randomly between states according to a local set of rules;
we then observe a finite collection of these sites at only a few times,
Suppose that the dynamics are Markov:
with the set of possible states by $\calS$,
and $X_i(t)$ the state of site $i$ at time $t$,
we assume that $\{X(t)\}_{t \ge 0}$ is a Markov process on sequences of $n$ states $\calS^n$
for which the probability a given site changes state in a small amount of time
depends only on the sequence of states at nearby sites.

@@

The dynamics of our stochastic process are determined by the set of allowed transitions and associated rates:
given two patterns $u$ and $v$ of common length $h$,
saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that
\[
    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
Such a rule is defined by its ``transition triple'' $(\mu,u,v)$,
written as ``$u \to v$ at rate $\mu$''.
The set of all transition triples, $\calT$, determines the time evolution of this process.

\peter{INCLUDE SELECTION}

\begin{example}[TASEP]
  The \emph{Totally Asymmetric Simple Exclusion Process} labels each site as ``empty'' or ``occupied'' (0 or 1, respectively),
  and says that each occupied site, independently at rate $\lambda$, checks the site to the right to see if it is empty,
  and if it is, moves there.
  This process therefore has only one transition triple:

@@

\begin{example}[1-D Ising model with Glauber dynamics]
  In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
  imagined as magnetic dipoles,
  and the energy associated with a given state $x$ is $H(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
  Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
  The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
  The following ``Glauber dynamics'' preserve the stationary distribution:
  each site, independently at rate $\lambda$,
  forgets its spin,
  and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration.
  Ignoring transitions that don't change the state,
  if patterns $u$ and $v$ differ at only one site then $u$ changes to $v$ at rate $\lambda/(1+\exp(H(v)-H(u)))$,
  so

    \peter{reformulate with a potential}

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$ per site,
and then resolving each possible change in temporal order.
The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
Note that since mutation is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
so the maximization over $i$ is unnecessary.
Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
so that for each site $i$,
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a draw from a rate $\mu_*$ Poisson process on $[0,t]$.
These processes are independent across sites.
Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $c_{ij} = \min_{k,h} \{ c_{kh} : c_{kh}>s \}$.
At $c_{ij}$,
a new state is then chosen at position $i$
with probability proportional to the corresponding rate, i.e.\
\erick{Previously, the $j$ in $\mu^j$ was used to index the possible patterns. Now it's indexing possible changes at a site.}
\[
q_j = \begin{cases}
  \mu_j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
is the probability that $x_i^{(|u^j|)}$ is replaced with $v^j$.
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$ gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0\ge 0$.)

@@

\peter{shorten, refer to figure}

@@

\peter{how to choose window size}

\peter{shorten/edit this to follow on from the above}

@@

The algorithm in general is written out in Appendix \peter{ref it}.

%%%%%%%
\section{Results}

\peter{from simulation: parameter estimation; model selection}

\peter{do a three-taxon tree? infer with shape data?}

The method works well to find point estimates of the parameters.
Obtaining posterior density estimates using the approximation likelihood function on a subset of patterns
works in some cases;
in other cases the likelihood surface is still too peaked, underestimating the uncertainty.
(but counting in nonoverlapping windows fixes the peakiness problem)

@@

\section{Discussion}

\begin{itemize}

@@

@@

  \item noisy image transmission denoising: first learn model; then reduce redundancy

  \item root distribution and confounding/mixing

  \item shrinkage prior on highly parameterized model

@@

\subsection{Acknowledgements}
Erick, Jessica, Graham, Yaniv, Sergey, Simon, Matt, Rasmus

\appendix

%%%%
\section*{Proof of the approximation}

\section*{Peeling algorithm}

@@

@@

%%%%
\section{Other likelihood functions}

\subsection{Summary statistics}

\section{Removed bits that we probably don't want}

The choice to treat long sequences as cyclical is generally of little importance.
However, we have defined in \eqref{eqn:G_defn} the basic building block of our approximation, $G(m)$,
to be the transition matrix \emph{for a circular sequence} of length $m$, where $m$ may be short.
\peter{no we haven't}
This seems strange -- but, what else would we do at the edges?
We want to use $G(\ell+m+r)$ to compute the chance that a given $(\ell+m+r)$-sequence ends up with another given $m$-sequence in the middle;
and if $\ell$ and $r$ are big enough relative to $t$, this will be a good approximation regardless of what we do at the edges.
One answer would be to not allow patterns hanging off of either end to match,
resulting in a slight underestimate of the chance of change.
A better approximation would be to use the marginal frequencies in some way to average over surroudning sequence
-- but, this requires marginal frequencies to be known, and these are not stationary.
However, the leftmost site in a randomly chosen window is a sample from the marginal distribution,
and if sites separated by distance $\ell+m+r$ are sufficiently uncorrelated,
wrapping the sequence around is a good approximation to averaging over marginal frequencies in the surrounding sequence.
It is easy to imagine situations where this would not be the best solution --
but again, if $\ell$ and $r$ are large enough, it won't matter much.
